Trees

### 2019-01-16 Trees_making.Rmd
removed update info
improved exercises after input from Alfredo and Nick

---
title: "Trees"
author: "ML"
date: "10 January 2019"
output: html_document
---

@@

Prepared by Malgorzata (Losia) Lagisz, BEES, UNSW, AU, losialagisz@yahoo.com.  
Last updated: 1 June 2016 by Losia Lagisz

@@

Ecological and evolutionary meta-analyses usually aggregate data acros multiple species (there are single-species or non-species meta analyses too, but these are  minority).
Since species are related due to shared evolutionary history, effect sizes of closely related species can be more similar to these of less-related specis. 
This kind of non-independence can be represented as a correlation matrix of phylogenetic relatedness and added as a random factor to multilevel meta-analytic models. 
To create a relatedness matrix among species we need a phylogenetic tree for the list of species from our data set.

Currently, the easiest way to build a custom phylogenetic tree usable for meta-analysis is by using an R package called *rotl*.

A few things to note:  
1. We can use a tree with or without branch lengths (divergence measures).  
2. The tree has to be binary (strictly bifurcating = no polytomies, i.e. only two branches are coming from each node).  
3. The tree needs to be rooted (has a node representing the most recent common ancestor of all the species in the tree).   
4. The tree tip labels (representing species) must match **exactly** with the species names in the data set (otherwise MA models will not run with them).  
5. Trees can be also made and used for higher taxonomic levels than species, e.g. genus - level (not covered in this tutorial, but general principles are the same).

Start from installing (as required) and loading R packages.

```{r prepare, message = FALSE}
#install.packages("devtools","fulltxt","ape","metafor")
#update.packages("fulltxt")
#devtools::install_github("ropensci/rotl", dependencies = TRUE, build_vignette=TRUE)
#install.packages("rotl", "treebase", "diagram", "dplyr")

library(readxl)
library(ape)
library(fulltext)
library(metafor)
library(rotl)
library(treebase)
library(diagram)
library(dplyr)
```

@@

We will try to create a simple tree using 10 commonly used taxa (lab model species) from across the Tree of Life.  
Rotl is an R package (https://peerj.com/preprints/1471/) allowing us to access synthetic phylogenetic tree available at Open Tree of Life database (https://opentree.wikispaces.com/).

@@

When we are happy with how the taxon names are matched to the Open Tree identifiers, we can pass them to the
function tol_induced_subtree to retrieve the relationships among these taxa (a trimmed sub-tree from the synthetic phylogenetic tree).  
The tree is returned as an ape::phylo object and it can be manipulated, printed and saved easily using functions from the ape package.

```{r rotl 10species plot}
tree <- tol_induced_subtree(ott_ids = taxa[["ott_id"]], label_format = "name")
plot(tree, cex=.8, label.offset =.1, no.margin = TRUE)
```

Note that the species names have the Open Tree identifiers appended to them and *E.coli* was replaced with "Emrcaott616ott617", which is non-taxon node within Bacteria. Given we dont have any ither bacteria in the dataset, and the position on the tree is correct, thats fine, we just need to rename the node back.

```{r rotl 10species names}
tree$tip.label #check tip labels
ecoli <- tol_node_info(ott_id=474506, include_lineage=TRUE)
ecoli
tax_lineage(ecoli)
tnrs_match_names(names = ("Escherichia")) #we can see what info is available for this genus - not much

#just fix the tip label
tree$tip.label <- gsub("mrcaott616ott617", "Escherichia coli", tree$tip.label) #replace the tip label back with our species name
tree$tip.labe
tree$tip.label <- gsub("_"," ", tree$tip.label) #get rid of the underscores
```

@@

```{r rotl 10species tree}
#see the tree info
tree
#tree$node.label <- NULL #you can delete internal node labels
plot(tree) #plot the tree
```

Do a final check whether the tri is binary and if the tip labels of the tree match exactly our (fixed for typos) list of species.

@@

#check the overlap with our species list
###check the overlap between species name from the data fila nad the tree file
intersect(as.character(tree$tip.label), myspecies_fixed) #10 matching names
setdiff(myspecies_fixed, as.character(tree$tip.label)) #0 mismatches
setdiff(as.character(tree$tip.label), myspecies_fixed) #0 mismatches
```

@@

### EXAMPLE 2 - using fixed supertree from a file

Create a phylogenetic tree for 65 bird species used in Rutkowska, J., Dubiec, A. & Nakagawa, S. (2014) All eggs are made equal: meta-analysis of egg sexual size dimorphism in birds. Journal of Evolutionary Biology. 27: 153-160 http://onlinelibrary.wiley.com/doi/10.1111/jeb.12282/abstract

There are a few main supertrees for birds, so they can be used to create a subtree containing only the species from the list.  
Here we will use on of these supertrees "Ericson backbone" downloadable from http://birdtree.org/ (Jetz, W., G. H. Thomas, J. B. Joy, K. Hartmann, and A. O. Mooers. 2012. The global diversity of birds in space and time. Nature 491:444-448.). The website http://birdtree.org/ has an online tool for subsetting their supertrees generating distribution of trees rather than one tree. For simplicity and demonstration puproses we will use R to subset (prune) a single bird supertree down to a smaller list of species.

@@

```{r list birds load} 
birds <- unique(read.csv("bird_list.csv")$species_name) #load our list of bird species and make the names unique
#str(birds)
birds_stree <- read.tree("Ericson.tre") #load bird supertree
birds_stree #9993 tips = species
str(birds_stree) #has edge (branch) lengths
birds_stree <- collapse.singles(birds_stree) #clean up singularities, just in case

###check the overlap between species name from our list  and the bird supertree
intersect(as.character(birds_stree$tip.label), birds) #51 - all our species matching - they are contained in the tree
setdiff(birds, as.character(birds_stree$tip.label)) #0 mismatches - just making sure we have all our species included in the supertree
```

@@

```{r list birds prune} 
pruned_birds_stree <- drop.tip(birds_stree,birds_stree$tip.label[-match(birds, birds_stree$tip.label)]) #pruning
#check the pruned tree
is.binary.tree(pruned_birds_stree) #TRUE
is.ultrametric(pruned_birds_stree) #TRUE

intersect(as.character(pruned_birds_stree$tip.label), birds) #51 - all our species matching - they are contained in the tree
setdiff(birds, as.character(pruned_birds_stree$tip.label)) #0 mismatches - just making sure we have all our species included in the tree

#write.tree(pruned_birds_stree, file = "birds_51sp_tree.tre", append = FALSE, digits = 10, tree.names = FALSE) #save, if needed
```

@@

```{r plot tree birds, fig.width=10, fig.height=10, message=FALSE}
plot(pruned_birds_stree, cex=0.8) #plots with branch lengths when available
```

@@

We will now handle a larger set of more closely related taxa, from a real data set.  
The data set we use the dataset *dat.curtis1998* (tudies on the Effects of Elevated CO2 Levels on Woody Plant Mass) included in the *metafor* package.

```{r load Curtis1998 data}
dat <- get(data(dat.curtis1998))
str(dat) # 102 rows, many columns including various data
```

@@

```{r create Curtis1998 species list}
species_Curtis1998 <- paste(dat$genus, dat$species, sep=" ") 
str(species_Curtis1998)
species_Curtis1998 <- unique(tolower(species_Curtis1998)) #make species name unique
length(species_Curtis1998) #37
species_Curtis1998 #note: "populusx euramericana" should be same as "populus euramericana"
species_Curtis1998 <- gsub("populusx euramericana", "populus euramericana", species_Curtis1998) #merge 2 partially overlapping names (typo?)
species_Curtis1998 <- unique(tolower(species_Curtis1998))
length(species_Curtis1998) #36
#capitalise first letters
species_Curtis1998 <- paste(toupper(substr(species_Curtis1998, 1, 1)), substr(species_Curtis1998, 2, nchar(species_Curtis1998)), sep="")
species_Curtis1998
```

We have 36 unique species_Curtis1998 names. We try to run rotl searc with them.

```{r match Curtis1998 species_Curtis1998 list to otl taxa list}
taxa <- tnrs_match_names(names = species_Curtis1998)
dim(taxa) #36
taxa
```

Note that some alternative names were found. We asssume these are correct ones for now.  
Now that the taxon names are matched to the Open Tree identifiers, we can pass them to the function tol_induced_subtree to retrieve the relationships among these taxa.

```{r get Curtis1998 tree}
tree <- tol_induced_subtree(ott_ids= taxa[["ott_id"]], label_format = "name") 
#The following OTT ids were not found: [3915043].
taxa[taxa$ott_id==3915043,] #populus euramericana Populus x canadensis             FALSE 3915043       TRUE HYBRID
#note: Populus × canadensis, known as Canadian poplar or Carolina poplar, is a naturally occurring hybrid of Populus nigra and Populus deltoides
```

We run into a problem with Populus - need to fix this by using an alternative name.

```{r fix Curtis1998 species_Curtis1998 list}
populus <- tnrs_match_names(names = ("populus deltoides"))
populus
#replace populus euramericana with populus deltoides (note 2 versions of the name)
species_Curtis1998 <- gsub("Populus euramericana", "Populus deltoides", species_Curtis1998)
```

```{r get Curtis1998 tree again}
taxa <- tnrs_match_names(names = species_Curtis1998)
head(taxa)
dim(taxa) #36
tree <- tol_induced_subtree(ott_ids= taxa[["ott_id"]], label_format = "name") 
str(tree)
```

@@

```{r check tree Curtis1998}
tree$tip.label
tree$tip.label <- gsub("_"," ", tree$tip.label) #get rid of underscores
#tree$node.label <- NULL #delete node labels
```

Replace back "populus euramericana"" with "populus deltoides"" and save the tree in the Newick format.

```{r replace back populus hybrid }
tree$tip.label <- gsub( "Populus deltoides", "Populus euramericana", tree$tip.label)

#check the tree match
intersect(as.character(tree$tip.label), species_Curtis1998) #matches - they are contained in the tree and our species list
setdiff(species_Curtis1998, as.character(tree$tip.label)) #2 mismatches - these are only in species_Curtis1998
setdiff(as.character(tree$tip.label), species_Curtis1998) #2 mismatches - these are only in tree$tip.label

#amend the tip labels on the tree to match the species names from Curtis1998
tree$tip.label <- gsub("Fuscospora fusca", "Nothofagus fusca", tree$tip.label) #replace the tip label back with our species name
tree$tip.label <- gsub("Populus euramericana", "Populus deltoides", tree$tip.label) #replace the tip label back with our species name

#check the tree match
intersect(as.character(tree$tip.label), species_Curtis1998) #matches - they are contained in the tree and our species list
setdiff(species_Curtis1998, as.character(tree$tip.label)) #0 mismatches 
setdiff(as.character(tree$tip.label), species_Curtis1998) #0 mismatches

#write.tree(tree, file="tree_curtis1998.tre") #save the tree, if needed
#read.tree(file="tree_curtis1998.tre") #if you need to read it
```

@@

```{r check Curtis1998 tree binary}
is.binary.tree(tree) #FALSE (i.e. there are polytomies)
#is.ultrametric(tree) #the tree has no branch lengths, so cant be checked
```

@@

```{r resolve Curtis1998 polytomies at random}
set.seed(111) #making it replicable
tree_random <- multi2di(tree,random=TRUE)
is.binary.tree(tree_random) #TRUE
write.tree(tree_random, file="tree_curtis1998_random.tre") #save the tree
#read.tree(tree_random="tree_curtis1998_random.tre") #if you need to read it
```

@@

```{r plot Curtis1998 tree_random, fig.width=10, fig.height=8, echo=TRUE, message=FALSE}
plot(tree_random)
```

@@

3. Resolving manually by useing information from known more specialised published phylogeny without polytomies (see next exercise for how to find these).

@@

Finding trees using Treebase database and *treebase* package (Treebase package tutorial: https://cran.r-project.org/web/packages/treebase/vignettes/vignette.html and https://www.carlboettiger.info/2011/05/16/treebase-in-r-a-first-tutorial.html).  
We will try to find more resolved phylogeny for the four species from genus Acer that were included in Curtis1998 tree.
You can search Treebase directly at www.treebase.org, but we will do this from R (note both approaches can be quite slow).

We need to creat a custom search string with our four Acer species names from Curtis1998 dataset.

```{r make treebase searchstring}
#Acer <- search_treebase("Acer_saccharinumc", by="taxon", max_trees=1) #example of using with a single species name - note different commas than in the tutorial and the underscores!
Acer4sp <- species[grep("Acer ", species)]
Acer4sp
Acer4sp <- gsub(" ","_", Acer4sp) #get underscores back
Acer4sp_string <- paste(Acer4sp, sep = "", collapse = ' and ') 
```

Run the treebase search - get amx 50 trees that contain exactly these species names.

```{r search treebase}
Acer_treebase <- search_treebase(Acer4sp_string, by=c("taxon","taxon","taxon","taxon"), max_trees=50, exact_match = TRUE, only_metadata = TRUE) #note different commas than in the tutorial!
Acer_treebase
str(Acer_treebase)
#for easier viewing put these inn a data frame (note that multiple tree IDs can come from the same study).
t(data.frame(matrix(unlist(Acer_treebase), nrow=6, byrow=F), row.names = c("S.id", "Tr.id", "type", "kind", "quality", "ntax"))) #note the study IDs that you are using - you can find these on Treebase
```
There are multiple trees available from several studies.  
Unfortunately the functions in treebase that are supposed to fetch metadata for the trees dont seem to work (show_metadata, metadata - if you get them workinng please let me know).

From oooking at Treebase website I know that this one will be good:   
tree Tr418 Study S342 - Ackerly D., Donoghue M.J.	1998	Leaf size, sapling allometry, and Corner's rules: phylogeny and correlated evolution in maples (Acer).	American Naturalist

```{r dwonload single tree from treebase}
Acers_tree <- search_treebase("418", by="id.tree", exact_match = TRUE) #returns a single tree
Acers_tree <- Acers_tree[[1]]
```

Plot the tree dwonloaded from Treebase - looks good.

```{r plot Acers_tree, fig.width=10, fig.height=8, echo=TRUE, message=FALSE}
plot(Acers_tree)
```

@@

```{r plot pruned Acers_tree, fig.width=10, fig.height=4, echo=TRUE, message=FALSE}
plot(Acers4sp_tree)
```

@@

### EXAMPLE 5 - try it yourself

Use rotl to create a tree for the bunch of reptiles.   
*Remember*: Make sure that species names match exactly tree tip labels

@@

```{r rotl solution, echo = TRUE, eval = TRUE}
taxa2 <- tnrs_match_names(names = myspecies2)
taxa2 #visual check

tree2 <- tol_induced_subtree(ott_ids= taxa2[["ott_id"]])
plot(tree2, cex=.8, label.offset =.1, no.margin = TRUE)

#change name on the tip lebel to match the original species list
tree2$tip.label <- gsub("\\_ott.*","", tree2$tip.label)
tree2$tip.label <- gsub("_"," ", tree2$tip.label)
tree2$tip.label
#tree$node.label <- NULL #delete node labels

#check if the tree2 is really binary 
is.binary.tree(tree2) #TRUE (i.e. there are no polytomies)

#check the overlap between species name from the list nad the tree2 
intersect(as.character(tree2$tip.label), myspecies2) #12 matching names
setdiff(myspecies2, as.character(tree2$tip.label)) #2 mismatches - these are only in myspecies2
setdiff(as.character(tree2$tip.label), myspecies2) #2 mismatches - these are only in the tree2$tip.label

#rename the tip labels on the tree to match the species names from our list
tree2$tip.label <- gsub("Centrochelys sulcata", "Geochelone sulcata", tree2$tip.label) 
tree2$tip.label <- gsub("Iberolacerta cyreni", "Iberolacerta cyeni", tree2$tip.label) 
#Note: you should actually check which name from each pair is "more correct" and accordingly check the names either on the tree or in your species list / data set

#re-check the overlap between species name from the list nad the tree2 
intersect(as.character(tree2$tip.label), myspecies2) #14 matching names
setdiff(myspecies2, as.character(tree2$tip.label)) #0 mismatches - these are only in myspecies2
setdiff(as.character(tree2$tip.label), myspecies2) #0 mismatches - these are only in the tree2$tip.label
```

@@

### DISCUSSION

>>>>

---
title: "Trees"
author: "ML"
date: "16 January 2019"
output: html_document
---

@@

Prepared by Malgorzata (Losia) Lagisz, BEES, UNSW, AU, losialagisz@gmail.com

@@

Ecological and evolutionary meta-analyses usually aggregate data across multiple species (there are single-species or non-species meta analyses too, but these are  minority).
Since species are related due to shared evolutionary history, effect sizes of closely related species can be more similar to these of less-related species. 
This kind of non-independence can be represented as a correlation matrix of phylogenetic relatedness and added as a random factor to multilevel meta-analytic models. 
To create a relatedness matrix among species we need a phylogenetic tree for the list of species from our data set.

A few things to note:  
1. We can use a tree with or without branch lengths (divergence measures).  
2. The tree has to be binary (strictly bifurcating = no polytomies, i.e. only two branches are coming from each node).  
3. The tree needs to be rooted (has a node representing the most recent common ancestor of all the species in the tree).   
4. The tree tip labels (representing species) must match **exactly** with the species names in the data set (otherwise MA models will not run with them).  
5. Trees can be also made and used for higher taxonomic levels than species, e.g. genus-level (not covered in this tutorial, but general principles are the same).

The general flow of tree-making work is shown in the diagram below.  
You will usually start with species list from the mata-analytic data set (use binomial latin names for the species list and in your database). First, check if there is any phylogenetic tree available that already contains all your species (e.g. a supertree). Currently, the easiest way to build a custom phylogenetic tree usable for meta-analysis is by using an R package called *rotl*, which accesses a synthetic supertree from Open Tree of Life database (https://opentree.wikispaces.com/). We will use this approach in our examples below. Most of the time *rotl* will work well, especially with the common species. There are other super-trees available, e.g. for birds - we will use one of thes as our example too. Overall, if you can find a tree that contains all your specis, you just trim (subset) it down to your species. If a few species is missing, it is sometimes possible to substitute them with closely related species. If no existing phylogeny is readily usable, there are a few non-exclusive options depending on the range of species in the data set:  
 - If the species are very diverged (e.g. mixture of vertabrates and invertebrates), use NCBI Taxonomy Common Tree tool (http://www.ncbi.nlm.nih.gov/Taxonomy/CommonTree/wwwcmt.cgi). At the webpage, use an input box at the top of the page to search each species name or upload lst of the species names from a file. This allows to generate a scaffold for a tree.  
 - If few major clades are present, assemble together sub-trees created from bigger trees, if these exist for these clades.    
 - Use phyloGenerator (http://willpearse.github.io/phyloGenerator/) or other phylogenetic software to create phylogeny based on DNA sequences, if a shared sequence set exists for a given set of species (check if the resulting phylogeny makes sense).

If the resulting tree contains polytomies, these can be resolved at random, or based on information from more detailed trees (e.g. published phylogeneies of specific lower taxa). You can also graft missing species on your tree. Be aware of species names problems: typos, scientific name synonyms, subspecies - have to be re-checked / corrected / substituted if generate errors at any stage. Again, your final tree, must contain exactly the same list of species names as your meta-analytic data set.

```{r prepare, message = FALSE, echo = TRUE, eval = FALSE, warning = FALSE}
install.packages("diagram")
install.packages("tidyverse")
install.packages("ape")
install.packages("curl")
install.packages("fulltext")
install.packages("treebase")
install.packages("devtools")
devtools::install_github("ropensci/rotl", dependencies = TRUE, build_vignette=TRUE)
install.packages("rotl")
```

```{r load packages, message = FALSE, echo = TRUE, eval = TRUE, warning = FALSE}
library(diagram, tidyverse)
library(ape, curl)
library(fulltext, metafor)
library(treebase, devtools)
library(rotl)
```

Examples and exercises
-------

We will give you some examples on how to use *rotl* package to download sub-trees from the Open Tree of Life supertree, resolve polytomies, trim and plot trees and deal with some of the most common issues. We will try to do most of theses tasks in R, but doing many of these outside R (using websites, text and tree ditors, etc.) is also possible. We hope you get the idea of the main principles of the process.

Now, we start from installing (as required) and loading R packages we will use.

@@

We will try to create a simple tree using 10 commonly used taxa (lab model species) from across the tree of life.  
Rotl is an R package (https://peerj.com/preprints/1471/) allowing us to access synthetic phylogenetic tree available at Open Tree of Life database (https://opentree.wikispaces.com/).

@@

When we are happy with the taxa found Open Tree identifiers, we can pass them to the
function tol_induced_subtree to retrieve the relationships among these taxa (a trimmed sub-tree from the synthetic phylogenetic tree).

```{r rotl 10species tree, warning = FALSE}
tree <- tol_induced_subtree(ott_ids = taxa[["ott_id"]], label_format = "name") #we can get a Warning meassage here about collapsing single nodes
```

The tree is returned as an ape::phylo object and it can be manipulated, printed and saved easily using functions from the ape package.

```{r rotl 10species plot, fig.width=10, fig.height=5, echo=TRUE, message=FALSE}
plot(tree, cex=.8, label.offset =.1, no.margin = TRUE)
```

Note that the species names have the Open Tree identifiers appended to them and *E.coli* was replaced with "mrcaott616ott617", which is non-taxon node within Bacteria. Given we dont have any other bacteria in the dataset, and the position on the tree is correct, thats fine, we just need to rename the tree tip back to "Escherichia coli".

```{r rotl 10species names}
tree$tip.label #see the current tree tip labels
ecoli <- tol_node_info(ott_id=474506, include_lineage=TRUE)
tax_lineage(ecoli) #domain within Bacteria
tnrs_match_names(names = ("Escherichia")) #we can see what info is available for this genus - not much, ooks like a messy bit of a tree flagged as "INCONSISTENT"
tree$tip.label <- gsub("mrcaott616ott617", "Escherichia coli", tree$tip.label) #replace the tip label back with our species name
tree$tip.label <- gsub("_"," ", tree$tip.label) #get rid of the underscores
```

@@

```{r rotl 10species tree Ecoli, fig.width=10, fig.height=5, echo=TRUE, message=FALSE}
# tree #see the tree info
# tree$node.label <- NULL #you can delete internal node labels
plot(tree, cex=.8, label.offset =.1, no.margin = TRUE) #plot the tree
```

Do a final check whether the tree is binary and if the tip labels of the tree match exactly our (fixed for typos) list of species.

@@

#check the overlap between the species names from the data set (our species list) nad the final tree
intersect(as.character(tree$tip.label), myspecies_fixed) #10 matching names
setdiff(myspecies_fixed, as.character(tree$tip.label)) #0 mismatches
setdiff(as.character(tree$tip.label), myspecies_fixed) #0 mismatches
```

That was hopefully easy! We can move to a different set of species and method...

@@

### EXAMPLE 2 - using a supertree from a file

We will now create a phylogenetic tree for 65 bird species used in Rutkowska, J., Dubiec, A. & Nakagawa, S. (2014) All eggs are made equal: meta-analysis of egg sexual size dimorphism in birds. Journal of Evolutionary Biology. 27: 153-160 http://onlinelibrary.wiley.com/doi/10.1111/jeb.12282/abstract

There are a few main existing and established supertrees for birds, so they can be used to create a subtree containing only the species from the list.  
Here we will use on of these supertrees called "Ericson backbone", which is downloadable from http://birdtree.org/ (Jetz, W., G. H. Thomas, J. B. Joy, K. Hartmann, and A. O. Mooers. 2012. The global diversity of birds in space and time. Nature 491:444-448.). The website http://birdtree.org/ also has an online tool for subsetting their supertrees generating distribution of trees rather than one tree. For simplicity and demonstration puproses we will use R to subset (prune) a single bird supertree down to a smaller list of species.

@@

```{r list birds load} 
birds <- unique(read.csv("bird_list.csv")$species_name) #load our list of bird species and make the names unique
birds_stree <- read.tree("Ericson.tre") #load bird supertree
birds_stree #9993 tips = species
# str(birds_stree) # ypu can see this tree has edge (branch) lengths

###check the overlap between species name from our list  and the bird supertree
intersect(as.character(birds_stree$tip.label), birds) #51 - all our species matching - they are contained in the tree
setdiff(birds, as.character(birds_stree$tip.label)) #0 mismatches - just making sure we have all our species included in the supertree, if you get any mismatches here look for typos or name synonyms!
```

@@

```{r list birds prune} 
pruned_birds_stree <- drop.tip(birds_stree,birds_stree$tip.label[-match(birds, birds_stree$tip.label)]) #pruning
#check the pruned tree
is.binary.tree(pruned_birds_stree) #TRUE
is.ultrametric(pruned_birds_stree) #TRUE
has.singles(pruned_birds_stree) #check for singles - the single nodes (i.e., with a single descendant) in a tree
# pruned_birds_stree <- collapse.singles(pruned_birds_stree) #if singles are preset, you can clean them up using collapse.singles function

# intersect(as.character(pruned_birds_stree$tip.label), birds) #51 - all our species matching - they are contained in the tree
setdiff(birds, as.character(pruned_birds_stree$tip.label)) #0 mismatches - just making sure we have all our species included in the tree
# write.tree(pruned_birds_stree, file = "birds_51sp_tree.tre", append = FALSE, digits = 10, tree.names = FALSE) #save, if needed
```

@@

```{r plot tree birds, fig.width=10, fig.height=8, message=FALSE}
plot(pruned_birds_stree, cex=0.8) #plots with branch lengths when available
```

@@

We will now handle a larger set of more closely related taxa, from a real data set.  
The data set we use the dataset *dat.curtis1998* (contains studies on the Effects of Elevated CO2 Levels on Woody Plant Mass) included in the *metafor* package.

```{r load Curtis1998 data}
dat <- metafor::dat.curtis1998
str(dat) #102 rows, many columns including various data
```

@@

```{r create Curtis1998 species list}
species_Curtis1998 <- paste(dat$genus, dat$species, sep=" ") 
# str(species_Curtis1998)
species_Curtis1998 <- unique(tolower(species_Curtis1998)) #make species name unique
length(species_Curtis1998) #37
species_Curtis1998 #note: "populusx euramericana" should be same as "populus euramericana"
species_Curtis1998 <- gsub("populusx euramericana", "populus euramericana", species_Curtis1998) #merge 2 partially overlapping names (typo? or hybrid?) - remember to fix this in the main data set!
species_Curtis1998 <- unique(tolower(species_Curtis1998)) #make the names unique again
length(species_Curtis1998) #36
species_Curtis1998 <- paste(toupper(substr(species_Curtis1998, 1, 1)), substr(species_Curtis1998, 2, nchar(species_Curtis1998)), sep="") #capitalise first letters
species_Curtis1998 #looks better
```

We now have 36 unique species_Curtis1998 names. We try to run *rotl* search with them.

```{r match Curtis1998 species_Curtis1998 list to otl taxa list}
taxa <- tnrs_match_names(names = species_Curtis1998)
dim(taxa) #36 species
# taxa
```

If you display the names (taxa), note that some alternative names were found. We asssume these are correct ones for now.  
Now that the taxon names are matched to the Open Tree identifiers, we can pass them to the function tol_induced_subtree to retrieve the relationships among these taxa.

```{r get Curtis1998 tree}
tree <- tol_induced_subtree(ott_ids= taxa[["ott_id"]], label_format = "name") 
#We get an error message that one OTT id was not found: [3915043].
taxa[taxa$ott_id==3915043,] #populus euramericana Populus x canadensis             FALSE 3915043       TRUE HYBRID
#note: Populus × canadensis, known as Canadian poplar or Carolina poplar, is a naturally occurring hybrid of Populus nigra and Populus deltoides
```

So, we run into a problem with one of the Populus names - need to fix this by using an alternative name.

```{r fix Curtis1998 species_Curtis1998 list}
populus <- tnrs_match_names(names = ("populus deltoides"))
populus #looks ok
species_Curtis1998 <- gsub("Populus euramericana", "Populus deltoides", species_Curtis1998) #replace populus euramericana with populus deltoides
```

Re-run *rotl* tree search.

```{r get Curtis1998 tree again, warning = FALSE}
taxa <- tnrs_match_names(names = species_Curtis1998)
dim(taxa) #36
tree <- tol_induced_subtree(ott_ids= taxa[["ott_id"]], label_format = "name")  #we can get a Warning meassage here about collapsing single nodes
str(tree)
```

@@

Replace back "Populus euramericana" with "Populus deltoides" and save the tree in the Newick format.

```{r check replace back populus hybrid }
tree$tip.label <- gsub("_"," ", tree$tip.label) #get rid of underscores
# tree$node.label <- NULL #delete node labels

#check the tree match
# intersect(as.character(tree$tip.label), species_Curtis1998) #matches - they are contained in the tree and our species list
setdiff(species_Curtis1998, as.character(tree$tip.label)) #1 mismatch - these are only in species_Curtis1998
setdiff(as.character(tree$tip.label), species_Curtis1998) #1 mismatch - these are only in tree$tip.label

#amend the tip labels on the tree to match the species names from Curtis1998
tree$tip.label <- gsub("Fuscospora fusca", "Nothofagus fusca", tree$tip.label) #replace the tip label back with our species name

#check the tree match again
# intersect(as.character(tree$tip.label), species_Curtis1998) #matches - they are contained in the tree and our species list
setdiff(species_Curtis1998, as.character(tree$tip.label)) #0 mismatches 
setdiff(as.character(tree$tip.label), species_Curtis1998) #0 mismatches

# write.tree(tree, file="tree_curtis1998.tre") #save the tree in Newick format, if needed
# read.tree(file="tree_curtis1998.tre") #if you need to read it
```

@@

```{r check Curtis1998 tree binary}
is.binary.tree(tree) #FALSE (i.e. there are polytomies)
# is.ultrametric(tree) #the tree has no branch lengths, so cant be checked
```

@@

```{r resolve Curtis1998 polytomies at random}
set.seed(111) #making it replicable
tree_random <- multi2di(tree,random=TRUE)
is.binary.tree(tree_random) #TRUE
write.tree(tree_random, file="tree_curtis1998_random.tre") #save the tree
# read.tree(file="tree_curtis1998_random.tre") #if you need to read it
```

@@

```{r plot Curtis1998 tree_random, fig.width=10, fig.height=8, echo=TRUE, message=FALSE}
plot(tree_random, cex=.8, label.offset =.1, no.margin = TRUE)
```

@@

3. Resolving manually by useing information from known more specialised published phylogeny without polytomies (see next exercise for how to try to find these).

@@

We can try to find some relevant trees using Treebase database and *treebase* package (Treebase package tutorial: https://cran.r-project.org/web/packages/treebase/vignettes/vignette.html and https://www.carlboettiger.info/2011/05/16/treebase-in-r-a-first-tutorial.html).

As an exercise, we will try to find more resolved phylogeny for the four species from genus Acer that were included in Curtis1998 from R.
However, Treebase sometimes may not connect or run the search string correctly. You can try search Treebase directly at www.treebase.org, but sometimes its quite slow or not working.
So, just in case we also provide the full tree files downloaded from there, which can be used in this exercise. In R, we first try to find a tree via Treebase website search - go to https://treebase.org/treebase-web/search/studySearch.html and enter Acer in teh search box and select "Abstract" from the drop-down menu next to it, press "Search" button. There are 4 relevant studies which potentially contain relevant trees - we will use the S342 Ackerly D., Donoghue M.J.	1998, so check the box next to it and then "Discard unchecked items" button at the bottom. Then, go to the "Trees"" tab where you will se 2 trees from this study. We will use Tr418 - you can download a Nexus file by clicking the icons on the right. Not: you can start the search from Tree tab, by searching tree titles and then look at the studies from which these trees are, or even search for the most obscure species in the species tab - sometimes these approaches work better.

Now we will try to do similar work from R. First, we create a custom search string with our four Acer species names from Curtis1998 dataset.

```{r make treebase searchstring}
Acer4sp <- species_Curtis1998[grep("Acer ", species_Curtis1998)]
Acer4sp #you should get "Acer rubrum"        "Acer saccharinum"   "Acer pensylvanicum" "Acer saccharum"  
Acer4sp <- gsub(" ","_", Acer4sp) #put underscores back in the names
Acer4sp_string <- paste(Acer4sp, sep = "", collapse = ' and ') 
Acer4sp_string #you should get "Acer_rubrum and Acer_saccharinum and Acer_pensylvanicum and Acer_saccharum"
```

Run the treebase search - get amx 50 trees that contain exactly these four species names.

```{r search treebase}
#this is an example of how to search for a single species:
# Acer <- search_treebase("Acer_saccharinumc", by="taxon", max_trees=10) #example of using with a single species name - note different commas than in the tutorial and the underscores!

Acer_treebase <- search_treebase(Acer4sp_string, by=c("taxon","taxon","taxon","taxon"), max_trees=50, exact_match = TRUE, only_metadata = TRUE) #note different commas than in the treebase tutorial!
# Acer_treebase #list of 49 meta-data results of our search 
# str(Acer_treebase) #some more info can be seen here

#Unfortunately the functions in treebase that are supposed to fetch metadata for the trees dont seem to work (show_metadata, metadata - if you get them workinng please let me know). 
#For easier viewing we will put these inn a data frame (note that multiple tree IDs can come from the same study).
t(data.frame(matrix(unlist(Acer_treebase), nrow=6, byrow=F), row.names = c("S.id", "Tr.id", "type", "kind", "quality", "ntax"))) #note the study IDs that you are using - you can find these on Treebase
```

There are multiple trees available from several studies. Note that the lost of found results includes the tree 418 from study 342 we already identified in our Treebase website search.  
So, we will request specifically tree Tr418 Study S342 - Ackerly D., Donoghue M.J.	1998	Leaf size, sapling allometry, and Corner's rules: phylogeny and correlated evolution in maples (Acer).	American Naturalist. It may take a moment, depending on the size of the tree. You can try requesting different trees later.

```{r dwonload single tree from treebase}
Acers_tree <- search_treebase("418", by="id.tree", exact_match = TRUE) #returns a single tree in a list
Acers_tree <- Acers_tree[[1]] #extract tree from the list
```

If downloading from R treebase did not work, upload tree saved in a file:
```{r dwonload single tree from file, eval = FALSE}
Acers_tree <- read.nexus(file = "T418.nex") #read Nexus tree file
```

Plot the Acers tree from Treebase - looks good.

```{r plot Acers_tree, fig.width=10, fig.height=8, echo=TRUE, message=FALSE}
plot(Acers_tree, cex=.8, label.offset =.1, no.margin = TRUE)
```

@@

```{r plot pruned Acers_tree, fig.width=10, fig.height=4, echo=TRUE, message=FALSE}
plot(Acers4sp_tree, cex=.8, label.offset =.1, no.margin = TRUE)
```

@@

### EXAMPLE 5 - try it yourself - rotl

Use *rotl* to create a tree for the bunch of reptiles.   
**Remember**: Make sure that species names match exactly tree tip labels.

Here is the species list:

@@

@@

### EXAMPLE 6 - try it yourself- treebase

Use *treebase* to create a tree for the three Quercus species from dat.Curtis1998.

**Hint**: Quercus prinus is also called Quercus montana (https://en.wikipedia.org/wiki/Quercus_montana).   
Once you substitute, all three species can be found on Tree 113123 of Study 22057, and also in several other studies and trees.

Here is the species list:

```{r treebase exercise, echo = TRUE, eval = TRUE, warning = FALSE}
Quercus3sp <- species_Curtis1998[grep("Quercus ", species_Curtis1998)]
```

### DISCUSSION
W hope you learn a few tricks on how to make phylogenetic-like trees of species that can be used in meta-analytic models. There is much more to learn, especially dealing with missing or rare species names, grafting trees together, resolving polytomies, plotting cool trees etc. In some cases getting "perfect" tree may not be possible - but phylogenies are always just estimats of the species relationships, and as long as you get the key features right, the tree will be ok, and so your analyses (in most cases).
